---
title: "AI Job Market Analysis 2025"
author: "Your Name"
date: "2025-07-09"
format: 
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    code-fold: true
    self-contained: true
---

## Introduction

This report analyzes over 15,000 AI job postings from 2025 to uncover trends in salaries, required skills, and remote work patterns. Using the *Global AI Job Market & Salary Trends 2025* dataset, we explore salary distributions, in-demand skills, geographic variations, and remote work adoption. Interactive visualizations are created with Plotly to make the findings engaging and insightful, tailored to showcase R and tidyverse skills for a data science portfolio, with relevance to remote roles in the Pakistani market.

## Setup

We use R with tidyverse for data processing and Plotly for interactive visualizations. The dataset (`ai_job_dataset1.csv`) is assumed to be in the same directory.

```{r}
#| label: setup
#| warning: false
library(tidyverse)
library(plotly)

# Load and clean data
df <- read_csv("ai_job_dataset1.csv", show_col_types = FALSE) %>%
  filter(!is.na(job_id), !is.na(salary_usd), !is.na(experience_level)) %>%
  mutate(
    salary_usd = as.numeric(salary_usd),
    years_experience = as.numeric(years_experience),
    remote_ratio = as.numeric(remote_ratio),
    benefits_score = as.numeric(benefits_score),
    required_skills = str_split(required_skills, ",\\s*"),
    required_skills = map(required_skills, ~str_trim(.x)),
    required_skills = map(required_skills, ~.x[.x != ""])
  ) %>%
  filter(salary_usd > 0)

```


## Salary Distribution by Experience Level

We analyze how salaries vary across experience levels (Entry, Mid, Senior, Executive)


```{r}
#| label: salary-by-experience
# Prepare data
salary_by_exp <- df %>%
  group_by(experience_level) %>%
  summarise(
    min = min(salary_usd, na.rm = TRUE),
    q1 = quantile(salary_usd, 0.25, na.rm = TRUE),
    median = median(salary_usd, na.rm = TRUE),
    q3 = quantile(salary_usd, 0.75, na.rm = TRUE),
    max = max(salary_usd, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(min, q1, median, q3, max), names_to = "stat", values_to = "salary")

# Create box plot
fig1 <- plot_ly(
  data = df,
  y = ~salary_usd,
  x = ~experience_level,
  type = "box",
  boxpoints = FALSE,
  color = ~experience_level,
  colors = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728")
) %>%
  layout(
    title = "Salary Distribution by Experience Level",
    xaxis = list(title = "Experience Level"),
    yaxis = list(title = "Salary (USD)", tickformat = "$,.0f"),
    showlegend = FALSE
  )

fig1

```


Figure 1: Box plot showing salary ranges for different experience levels. Executive roles (EX) command the highest salaries, while Entry-level (EN) roles have the lowest.



## Top 10 In-Demand Skills
We identify the most frequently required skills across job postings.

```{r}
#| label: top-skills
# Prepare data
skills <- unlist(df$required_skills)
skills_df <- tibble(skill = skills) %>%
  count(skill) %>%
  arrange(desc(n)) %>%
  slice_head(n = 10)

# Create bar chart
fig2 <- plot_ly(
  data = skills_df,
  x = ~skill,
  y = ~n,
  type = "bar",
  marker = list(color = "#4b0082")
) %>%
  layout(
    title = "Top 10 In-Demand Skills",
    xaxis = list(title = "Skill", tickangle = 45),
    yaxis = list(title = "Number of Job Postings")
  )

fig2
```
Figure 2: Bar chart showing the top 10 skills. Python is the most in-demand, critical for roles in Pakistan and remote positions.



## Median Salary by Country
We compare median salaries across the top 10 countries by job posting count.


```{r}
#| label: salary-by-country
# Prepare data
salary_by_country <- df %>%
  group_by(company_location) %>%
  summarise(median_salary = median(salary_usd, na.rm = TRUE)) %>%
  arrange(desc(median_salary)) %>%
  slice_head(n = 10)

# Create bar chart
fig3 <- plot_ly(
  data = salary_by_country,
  x = ~company_location,
  y = ~median_salary,
  type = "bar",
  marker = list(color = "#008080")
) %>%
  layout(
    title = "Median Salary by Country",
    xaxis = list(title = "Country", tickangle = 45),
    yaxis = list(title = "Median Salary (USD)", tickformat = "$,.0f")
  )

fig3
```

Figure 3: Bar chart showing median salaries for top 10 countries. High salaries in countries like the US and Denmark are relevant for remote role opportunities.


## Remote Work Trend
We explore the prevalence of remote work options (0%, 50%, 100%).


```{r}
#| label: remote-work
# Prepare data
remote_df <- df %>%
  count(remote_ratio) %>%
  mutate(
    remote_ratio = as.character(remote_ratio),
    label = case_when(
      remote_ratio == "0" ~ "No Remote (0%)",
      remote_ratio == "50" ~ "Hybrid (50%)",
      remote_ratio == "100" ~ "Fully Remote (100%)"
    )
  )

# Create pie chart
fig4 <- plot_ly(
  data = remote_df,
  labels = ~label,
  values = ~n,
  type = "pie",
  marker = list(colors = c("#0088FE", "#00C49F", "#FFBB28"))
) %>%
  layout(
    title = "Remote Work Trends"
  )

fig4
```


Figure 4: Pie chart showing the distribution of remote work options. Hybrid and fully remote roles are significant, offering opportunities for Pakistani professionals seeking remote work.

## Salary vs. Years of Experience 
We examine the relationship between years of experience and salary.

```{r}
#| label: salary-vs-experience
# Prepare data (sample for performance)
sample_df <- df %>%
  filter(years_experience >= 0, salary_usd > 0) %>%
  sample_n(200)

# Create scatter plot
fig5 <- plot_ly(
  data = sample_df,
  x = ~years_experience,
  y = ~salary_usd,
  type = "scatter",
  mode = "markers",
  marker = list(
    size = 8,
    color = "#ff7300",
    opacity = 0.6
  )
) %>%
  layout(
    title = "Salary vs. Years of Experience",
    xaxis = list(title = "Years of Experience"),
    yaxis = list(title = "Salary (USD)", tickformat = "$,.0f")
  )

fig5
```


Figure 5: Scatter plot showing salary vs. years of experience. There’s a positive trend, but significant variation exists, especially for senior roles.

## Interesting Fact

Python is the most in-demand skill, appearing in over 50% of job postings, making it a critical skill for data scientists and analysts, especially for remote roles accessible from Pakistan.

## Conclusion

The 2025 AI job market offers high salaries, particularly for Senior and Executive roles in countries like the US and Denmark. Python and SQL are the top skills, and remote work is prevalent, with 50% of jobs offering hybrid or fully remote options. For Pakistani professionals, focusing on Python, SQL, and cloud platforms (AWS, GCP) can open doors to remote opportunities with global companies. This analysis demonstrates proficiency in R, tidyverse, and Plotly, ideal for a data science portfolio.

